Case Study: ChIP-seq Differential Peak Calling (csaw)

### 2018-04-04 datasets/ChIP-seq/ChIPseq-CBP-csaw.Rmd
Add first pass at CBP ChIPseq analysis

In this case study, we perform differential peak calling on ChIP-seq data of the histone
H3K4Me3 for samples from two cell lines (K562 and GM12878) using publicly available data
generated by the ENCODE Project. The same data set is used for all ChIP-seq differential
testing case studies.

The _csaw_ package provides a powerful and flexible approach to differential peak calling
based on counting reads in sliding windows across the genome. The sliding window approach
allows for the detection of differential peaks over sharper or wider regions, without being
constrained by the original peak boundaries obtained from the peak calling algorithm, e.g. _MACS_.
Differential peaks are called at each window using _edgeR_ and p-values are combined within
adjacent windows (clusters).

For this analysis, the default _csaw_ settings are used as described in the _csaw_ "User Guide."

@@

## intermediary files we create below
count_file <- file.path(resdir, "h3k4me3-csaw-counts.rds")
filtered_file <- file.path(resdir, "h3k4me3-csaw-counts-filtered.rds")
result_file <- file.path(resdir, "h3k4me3-csaw-results.rds")
bench_file <- file.path(resdir, "h3k4me3-csaw-benchmark.rds")

@@

We download the fastq files directly from the UCSC ENCODE portal.

```{r, download-fastqs}
broad_url <- "http://hgdownload.soe.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeBroadHistone/"
broad_fastqs <- c("wgEncodeBroadHistoneGm12878H3k4me3StdRawDataRep1.fastq.gz",
                  "wgEncodeBroadHistoneGm12878H3k04me3StdRawDataRep2V2.fastq.gz",
                  "wgEncodeBroadHistoneK562H3k4me3StdRawDataRep1.fastq.gz",
                  "wgEncodeBroadHistoneK562H3k4me3StdRawDataRep2.fastq.gz")

for (i_fq in broad_fastqs) {
    if (!file.exists(file.path(datdir, i_fq))) {
        download.file(paste0(broad_url, i_fq),
                      destfile = file.path(datdir, i_fq))
    }
}

uw_url <- "http://hgdownload.soe.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeUwHistone/"
uw_fastqs <- c("wgEncodeUwHistoneGm12878H3k4me3StdRawDataRep1.fastq.gz",
               "wgEncodeUwHistoneGm12878H3k4me3StdRawDataRep2.fastq.gz",
               "wgEncodeUwHistoneK562H3k4me3StdRawDataRep1.fastq.gz",
               "wgEncodeUwHistoneK562H3k4me3StdRawDataRep2.fastq.gz")

for (i_fq in uw_fastqs) {
    if (!file.exists(file.path(datdir, i_fq))) {
        download.file(paste0(uw_url, i_fq),
                      destfile = file.path(datdir, i_fq))
    }
}
```

We use the GRCh38 reference genome used in the ENCODE3 analysis. If the reference index is
not already available, we must first download the reference file. Note that this is a different
reference than what was originally used for the data in earlier analyses of the ENCODE project.

```{r, download-ref}
ref_url <- paste0("https://www.encodeproject.org/files/",
                  "GRCh38_no_alt_analysis_set_GCA_000001405.15/@@download/",
                  "GRCh38_no_alt_analysis_set_GCA_000001405.15.fasta.gz")
ref_fastagz <- file.path(datdir, basename(ref_url))
ref_fasta <- gsub("\\.gz$", "", ref_fastagz)

if (!file.exists(file.path(datdir, "ref_index.00.b.tab"))) {
    if (!file.exists(ref_fasta)) {
        download.file(ref_url, destfile = ref_fastagz)
        system(paste("gunzip", ref_fastagz))
    }
    buildindex(basename = file.path(datdir, "ref_index"),
               reference = ref_fasta)
}
```

We determine sample metadata from the file names.

```{r}
fqfiles <- file.path(datdir, c(broad_fastqs, uw_fastqs))
labs <- gsub("wgEncode(.*)Histone.*", "\\1", basename(fqfiles))
cells <- gsub("wgEncode.*Histone(.*)H3k.*", "\\1", basename(fqfiles))

meta <- data.frame(cellline = cells, lab = labs,
                   fqfile = fqfiles,
                   bamfile = gsub("\\.fastq.gz", ".bam", fqfiles),
                   sortedfile = gsub("\\.fastq.gz", ".sorted.bam", fqfiles),
                   stringsAsFactors = FALSE)
```

We also download blacklisted regions for GRCh38 from ENCODE (https://www.encodeproject.org/annotations/ENCSR636HFF/).

```{r}
blacklist_url <- paste0("http://mitra.stanford.edu/kundaje/akundaje/",
                        "release/blacklists/hg38-human/hg38.blacklist.bed.gz")
if (!file.exists(file.path(datdir, basename(blacklist_url)))) {
    download.file(blacklist_url, destfile = file.path(datdir, basename(blacklist_url)))
}

@@

```{r}
std_chr <- paste0("chr", c(1:22, "X", "Y"))
param <- readParam(minq = 20, discard = bl, restrict = std_chr)
```

@@

```{r count, results = 'hide', message = FALSE, warning = FALSE}
if (file.exists(count_file)) {
    win_cnts <- readRDS(count_file)
} else {
    ## Broad samples have quality score w/ phred offset +33 (standard)
    b_meta <- meta[meta$lab == "Broad", ]
    b_unaligned <- !file.exists(b_meta$bamfile)
    if (any(b_unaligned)) {
        align(index = file.path(datdir, "ref_index"),
              readfile1 = b_meta$fqfile[b_unaligned],
              TH1 = 2, type = 1, phredOffset = 33,
              output_file = b_meta$bamfile[b_unaligned])
    }

    ## UW samples have quality score w/ phred offset +64 (old Illumina scale)
    u_meta <- meta[meta$lab == "Uw", ]
    u_unaligned <- !file.exists(u_meta$bamfile)
    if (any(u_unaligned)) {
        align(index = file.path(datdir, "ref_index"),
              readfile1 = u_meta$fqfile[u_unaligned],
              TH1 = 2, type = 1, phredOffset = 64,
              output_file = u_meta$bamfile[u_unaligned])
    }

@@

    ## mark duplicates w/ picard
    temp_bam <- file.path(datdir, "temp_dups.bam")
    temp_file <- file.path(datdir, "temp_mets.txt")
    temp_dir <- file.path(datdir, "temp_dups")
    dir.create(temp_dir)
    for (i_bam in meta$sortedfile) {
        code <- paste0("java -jar ${PICARD_TOOLS_HOME}/picard.jar ",
                       "MarkDuplicates I=%s O=%s M=%s ", 
                       "TMP_DIR=%s AS=true REMOVE_DUPLICATES=false ",
                       "VALIDATION_STRINGENCY=SILENT")
        code <- sprintf(code, i_bam, temp_bam, temp_file, temp_dir)
        code <- system(code)
        stopifnot(code == 0L)
        file.rename(temp_bam, i_bam)
    }
    unlink(temp_file)
    unlink(temp_dir, recursive = TRUE)

@@

    ## use correlateReads to determine fragment length (remove dups)
    b_x <- correlateReads(b_meta$sortedfile, param = reform(param, dedup = TRUE))
    u_x <- correlateReads(u_meta$sortedfile, param = reform(param, dedup = TRUE))
    b_frag_len <- which.max(b_x) - 1
    u_frag_len <- which.max(u_x) - 1

    ## count reads in sliding windows (keep dups)
    win_cnts <- windowCounts(meta$sortedfile, param = param, width = 150,
                             ext = list(rep(c(b_frag_len, u_frag_len), each = 4), 150))

@@

```{r}
if (file.exists(filtered_file)) {
    filtered_cnts <- readRDS(filtered_file)
} else {
    ## filter windows by abundance
    bins <- windowCounts(meta$sortedfile, bin = TRUE, width = 2000, param = param)
    filter_stat <- filterWindows(win_cnts, bins, type = "global")
    keep <- filter_stat$filter > log2(2)
    filtered_cnts <- win_cnts[keep, ]

    ## save filtered counts
    saveRDS(filtered_cnts, file = filtered_file)
}
```

@@

```{r test, results='hide', message=FALSE, warning=FALSE}
if (file.exists(result_file)) {
    res_ranges <- readRDS(result_file)
} else {
    ## normalize for library-specific trended biases
    offsets <- normOffsets(filtered_cnts, type = "loess")

    ## create model design
    design <- model.matrix(~ lab + cellline,
                           data = as.data.frame(meta))

    ## run quasi-likelihood F-test
    y <- asDGEList(filtered_cnts)
    y$offset <- offsets
    y <- estimateDisp(y, design)
    fit <- glmQLFit(y, design, robust = TRUE)
    res <- glmQLFTest(fit)

@@

    ## save results
    res_ranges <- merged$region
    elementMetadata(res_ranges) <-
        data.frame(tab_comb,
                   best_pos = mid(ranges(rowRanges(filtered_cnts[tab_best$best]))),
                   best_logFC = tab_best$logFC)

    saveRDS(res_ranges, file = result_file)
}

>>>>

In this case study, we perform differential peak calling on ChIP-seq data for a
trnscription factor, CREB-binding protein (CBP), from Kasper et al. (2014), used in
Lun et al. (2015) to demonstrate the usage of _csaw_. As described in a separate case
study analyzing ChIP-seq data for H3K4Me3, _csaw_ is a package for differential peak
calling based on counting reads in sliding windows across the genome. Code and steps
for initial processing and analysis of this data with _csaw_ are taken directly
from Lun et al. (2015).

@@

## intermediary files we create below
count_file <- file.path(resdir, "cbp-csaw-counts.rds")
filtered_file <- file.path(resdir, "cbp-csaw-counts-filtered.rds")
normfacs_file <- file.path(resdir, "cbp-csaw-counts-normfacs.rds")
result_file <- file.path(resdir, "cbp-csaw-results.rds")
bench_file <- file.path(resdir, "cbp-csaw-benchmark.rds")

@@

We download the fastq files directly from the European Nucleotide Archive (ENA).

```{r download-fastqs}
fqurls <- c("007/SRR1145787/SRR1145787.fastq.gz",
            "008/SRR1145788/SRR1145788.fastq.gz",
            "009/SRR1145789/SRR1145789.fastq.gz",
            "000/SRR1145790/SRR1145790.fastq.gz")
fqurls <- paste0("ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR114/", fqurls)

for (i_fq in fqurls) {
    out_fq <- file.path(datdir, basename(i_fq))
    if (!file.exists(out_fq)) {
        download.file(i_fq, destfile = out_fq)
    }
}
```

We download the mm10 reference genome used for ENCODE3 and build the
index for alignment if not already available.

```{r, download-ref}
ref_url <- paste0("https://www.encodeproject.org/files/",
                  "mm10_no_alt_analysis_set_ENCODE/@@download/",
                  "mm10_no_alt_analysis_set_ENCODE.fasta.gz")
ref_fastagz <- file.path(datdir, basename(ref_url))
ref_fasta <- gsub("\\.gz$", "", ref_fastagz)

if (!file.exists(file.path(datdir, "mm_ref_index.00.b.tab"))) {
    if (!file.exists(ref_fasta)) {
        download.file(ref_url, destfile = ref_fastagz)
        system(paste("gunzip", ref_fastagz))
    }
    buildindex(basename = file.path(datdir, "mm_ref_index"),
               reference = ref_fasta)
}
```

Sample metadata is stored in a data.frame.

```{r}
fqfiles <- file.path(dardir, basename(fqurls)
meta <- data.frame(genotype = factor(c("wt", "wt", "ko", "ko")),
                   fqurl = fqurls, fqfile = fqfiles,
                   bamfile = gsub("\\.fastq\\.gz", ".bam", fqfiles),
                   sortedfile = gsub("\\.fastq\\.gz", ".sorted.bam", fqfiles))
```

We also download blacklisted regions for mm10 from ENCODE (https://www.encodeproject.org/annotations/ENCSR636HFF/).

```{r}
blacklist_url <- "https://www.encodeproject.org/files/ENCFF547MET/@@download/ENCFF547MET.bed.gz"
if (!file.exists(file.path(datdir, basename(blacklist_url)))) {
    download.file(blacklist_url, destfile = file.path(datdir, basename(blacklist_url)))
}

@@

```{r}
std_chr <- paste0("chr", c(1:19, "X", "Y"))
param <- readParam(minq = 20, discard = bl, restrict = std_chr)
```

@@

```{r count, results = 'hide', message = FALSE, warning = FALSE}
if (file.exists(count_file)) {
    win_cnts <- readRDS(count_file)
} else {
    unaligned <- !file.exists(meta$bamfile)
    if (any(unaligned)) {
        align(index = file.path(datdir, "mm_ref_index"),
              readfile1 = meta$fqfile[unaligned],
              type = 1, phredOffset = 64,
              input_format = "FASTQ",
              output_file = meta$bamfile[unaligned])
    }

@@

    ## mark duplicates w/ picard
    temp_bam <- file.path(datdir, "temp_dups_mm.bam")
    temp_file <- file.path(datdir, "temp_mets_mm.txt")
    temp_dir <- file.path(datdir, "temp_dups_mm")
    dir.create(temp_dir)
    for (i_bam in meta$sortedfile) {
        code <- paste0("java -jar ${PICARD_TOOLS_HOME}/picard.jar ",
                       "MarkDuplicates I=%s O=%s M=%s ", 
                       "TMP_DIR=%s AS=true REMOVE_DUPLICATES=false ",
                       "VALIDATION_STRINGENCY=SILENT")
        code <- sprintf(code, i_bam, temp_bam, temp_file, temp_dir)
        code <- system(code)
        stopifnot(code == 0L)
        file.rename(temp_bam, i_bam)
    }
    unlink(temp_file)
    unlink(temp_dir, recursive = TRUE)

@@

    ## use correlateReads to determine fragment length (remove dups)
    x <- correlateReads(b_meta$sortedfile, param = reform(param, dedup = TRUE))
    frag_len <- which.max(x) - 1

    ## count reads in sliding windows (keep dups)
    win_cnts <- windowCounts(meta$sortedfile, param = param, width = 10, ext = frag.len)

@@

```{r}
if (file.exists(filtered_file)) {
    filtered_cnts <- readRDS(filtered_file)
    normfacs <- readRDS(normfacs_file)
} else {
    ## filter windows by abundance
    bins <- windowCounts(meta$sortedfile, bin = TRUE, width = 10000, param = param)
    filter_stat <- filterWindows(win_cnts, bins, type = "global")
    keep <- filter_stat$filter > log2(3)
    filtered_cnts <- win.data[keep, ]

    ## calculate normalizing offsets based on larger windows
    normfacs <- normOffsets(bins)

    ## save filtered counts, normalizing factors
    saveRDS(filtered_cnts, file = filtered_file)
    saveRDS(normfacs, file = normfacs_file)
}
```

@@

```{r test, results='hide', message=FALSE, warning=FALSE}
if (file.exists(result_file)) {
    res_ranges <- readRDS(result_file)
} else {
    ## create model diesign
    design <- model.matrix(~ 0 + genotype, data = meta)
    colnames(design) <- levels(meta$genotype)

    ## run quasi-likelihood F-test
    y <- asDGEList(filtered_cnts, norm.factors = normfacs)
    y <- estimateDisp(y, design)
    fit <- glmQLFit(y, design, robust = TRUE)
    res <- glmQLFTest(fit, contrast = makeContrasts(wt-ko, levels = design))

@@

    ## save results
    res_ranges <- merged$region
    elementMetadata(res_ranges) <-
        data.frame(tabcom,
                   best_pos = mid(ranges(rowRanges(filtered_cnts[tab_best$best]))),
                   best_logFC = tab_best$logFC)
    saveRDS(res_ranges, file = result_file)
}
